An Introduction to Return Oriented Programming

### 2023-02-16 paper.tex
changes

%%%% 1. DOCUMENTCLASS %%%%
\documentclass[journal=tosc,submission]{iacrtrans}
%%%% NOTES:
% - Change "journal=tosc" to "journal=tches" if needed
% - Change "submission" to "final" for final version
% - Add "spthm" for LNCS-like theorems

@@

%%%% 3. AUTHOR, INSTITUTE %%%%
\author{Maximilian Heim}
\institute{
  University Albstadt-Sigmaringen, Albstadt, Germany, \email{MaximilianHeim@protonmail.com}
}
%%%% NOTES:
% - We need a city name for indexation purpose, even if it is redundant
%   (eg: University of Atlantis, Atlantis, Atlantis)
% - \inst{} can be omitted if there is a single institute,
%   or exactly one institute per author

%%%% 4. TITLE %%%%
\title{Return Oriented Programming}
%%%% NOTES:
% - If the title is too long, or includes special macro, please
%   provide a "running title" as optional argument: \title[Short]{Long}
% - You can provide an optional subtitle with \subtitle.

@@

\Verb+pop+ allows us to write arbitrary values into registers. For that we search for a \Verb+pop <reg>+ instruction inside our gadgets, in the payload we can then place the value that we want to insert after the address of the \Verb+pop+ instruction. If we can not find a suitable gadget we can try to get creative and achieve the desired state another way. For example if we want to modify \Verb+ecx+ but do not have a \Verb+pop ecx+ instruction available we could achieve it with something like this: \Verb+xor ecx, ecx ; pop eax ; xor ecx, eax+. Provided that we have these gadgets available.

@@

\Verb+mov+ allows us to write arbitrary values into memory. For that we search for a \Verb+mov dword ptr [<reg1>], <reg2>+ instruction inside our gadgets, we can then, in combination with two pops write arbitrary values at arbitary memory locations. The following example writes the value in \Verb+ecx+ to where \Verb+eax+ points to: \Verb+pop ecx ; pop eax ; mov dword ptr [eax], ecx+

@@

Arithmetic operations like \Verb+add+, \Verb+sub+, \Verb+inc+, \Verb+xor+, \Verb+or+, and can be useful to bring registers into our desired state. For that we search for the corresponding gadget with the required operands. For example \Verb+xor+ can be used to clear a register or copy its contents. It often occurs in the following forms: \Verb+xor eax, eax+ or \Verb+xor eax, edx+. The first case clears the register since \Verb+xor+ computes a non-equivalence, formally $a \oplus a = 0$ and the second one copies the value of the 2nd operand into the 1st operand when the target register is \Verb+0x00+ since \Verb+0x00+ is the neutral element of the \Verb+xor+ operation, formally $a \oplus 0 = a$.

@@

\section{Attack}
\label{sec:attack}
\subsection{Target Program}
\paragraph{Target Program}
The following program is the target of our attack, it uses a command line argument to provide the payload and \Verb+strcpy+ for the buffer overflow, overwriting the return address after the 8 Byte buffer. Using vulnerable input functions also works though.
\bltCode{vuln.c}{c}{The Target Program}{thetargetprogram}
\paragraph{Compilation}
We compile the target program with the following command. There are several important options given in this command. Most importantly the \Verb+-fno-stack-protector+ option disables stack canaries which would otherwise directly terminate the program when the canary is overwritten. The \Verb+-m32+ option compiles the binary as a 32 Bit executable, this makes the attack easier. The \Verb+-static+ option makes the binary statically linked. Without this option there are only 50 gadgets available, considering most of them are not useful for our attack it is practically impossible to perform the attack with just these gadgets. The \Verb+-static+ option includes the \Verb+libc+ library in the executable, increasing the gadget count to over 8000. However, it is possible to determine the address of the dynamically linked library at runtime and adding an offset for each gadget to this address. This has been described by Saif El-Sherei but will not be further discussed in this paper.
\bltCommand{compilation.sh}{The compliation command}{thecompilationcommand}
\subsection{Phases of developing the attack}
\paragraph{Phases}
The attack consists of several phases
\begin{enumerate}
  \item Specify concrete goal with required program state and instructions
  \item Generate desired list of instructions and arguments (abstract payload)
  \item Extract gadgets using tools
  \item Search gadgets for instructions
  \item Generate payload using the gadgets according to the the abstract payload while making sure gadgets dont interfere with our desired program state. This step can be done using Python which we will show in a later section~\cref{howtopack}
  \item Insert payload into target
\end{enumerate}
\paragraph{Goal and abstract payload}
After specifying the goal and possibly simplifying it we have to write a list of instructions and arguments that achieve the goal, for this its favorable to directly use the format of the final payload except for using instructions instead of addresses as this will then allow to simply insert the found gadgets into this abstract payload.
\paragraph{Extract and search gadgets}
After extracting the gadgets using one of the above mentioned methods we can search for gadgets
\paragraph{struct.pack}
\Verb+struct.pack+ is a Python function that allows to easily generate our desired payload from the raw bytes. Bash then allows to directly pipe the generated payload into our target. In order to generate the payload we first have to fill the buffer and override the EBP with arbitary values as seen in line 2~\cref{howtopack}. This is usually done using easily recognizable characters, using the letter \Verb+A+ for this is common. It has the hex value \Verb+0x41+, doing this allows then to spot the buffer in a debugger like \Verb+gdb+. So in this example we fill the buffer with 8 \Verb+A+'s and 4 \Verb+B+'s. After that it is time to insert the addresses of the gadgets and the arguments. This is done by calling pack with the double word (64 Bit) while specifying the endianness, converting that to a string and adding it to the string as seen in line 3~\cref{howtopack}. After the whole payload has been generated we can print it and use the output directly for running the buffer overflow attack as mentioned above.
\bltCode{pack.py}{python}{How to use struct.pack}{howtopack}

@@

\section{Discussion}
Sources:
\url{https://www.exploit-db.com/docs/english/28479-return-oriented-programming-(rop-ftw).pdf}
\url{https://guyinatuxedo.github.io/5.1-mitigation_aslr_pie/index.html}
%%%% 8. BILBIOGRAPHY %%%%
\bibliographystyle{alpha}
\bibliography{abbrev3,crypto,biblio}
%%%% NOTES
% - Download abbrev3.bib and crypto.bib from https://cryptobib.di.ens.fr/
% - Use bilbio.bib for additional references not in the cryptobib database.
%   If possible, take them from DBLP.
\end{document}

>>>>

%%%% 1. DOCUMENTCLASS %%%%
\documentclass[journal=tosc,submission, notanonymous]{iacrtrans}
%%%% NOTES:
% - Change "journal=tosc" to "journal=tches" if needed
% - Change "submission" to "final" for final version
% - Add "spthm" for LNCS-like theorems

@@

%%%% 3. AUTHOR, INSTITUTE %%%%
\author{Maximilian Heim\inst{1}}
\institute{
  University Albstadt-Sigmaringen, Albstadt, Germany, \email{MaximilianHeim@protonmail.com}
}
%%%% NOTES:
% - We need a city name for indexation purpose, even if it is redundant
%   (eg: University of Atlantis, Atlantis, Atlantis)
% - \inst{} can be omitted if there is a single institute,
%   or exactly one institute per author

%%%% 4. TITLE %%%%
\title{An Introduction to Return Oriented Programming}
%%%% NOTES:
% - If the title is too long, or includes special macro, please
%   provide a "running title" as optional argument: \title[Short]{Long}
% - You can provide an optional subtitle with \subtitle.

@@

\Verb+pop+ allows us to write arbitrary values into registers. For that we search for a \Verb+pop <reg>+ instruction inside our gadgets, in the payload we can then place the value that we want to insert after the address of the \Verb+pop+ instruction. If we can not find a suitable gadget we can try to get creative and achieve the desired state another way. For example if we want to modify \Verb+ecx+ but do not have a \Verb+pop ecx+ instruction available we could achieve it with something like this: \bltInlineVerb{xor ecx, ecx ; pop eax ; xor ecx, eax}. Provided that we have these gadgets available.

@@

\Verb+mov+ allows us to write arbitrary values into memory. For that we search for a \Verb+mov dword ptr [<reg1>], <reg2>+ instruction inside our gadgets, we can then, in combination with two pops write arbitrary values at arbitary memory locations. The following example writes the value in \Verb+ecx+ to where \Verb+eax+ points to: \bltInlineVerb{pop ecx ; pop eax ; mov dword ptr [eax], ecx}

@@

Arithmetic operations like \Verb+add+, \Verb+sub+, \Verb+inc+, \Verb+xor+, \Verb+or+, and can be useful to bring registers into our desired state. For that we search for the corresponding gadget with the required operands. For example \Verb+xor+ can be used to clear a register or copy its contents. It often occurs in the following forms: \bltInlineVerb{xor eax, eax} or \bltInlineVerb{xor eax, edx}. The first case clears the register since \Verb+xor+ computes a non-equivalence, formally $a \oplus a = 0$ and the second one copies the value of the 2nd operand into the 1st operand when the target register is \Verb+0x00+ since \Verb+0x00+ is the neutral element of the \Verb+xor+ operation, formally $a \oplus 0 = a$.

@@

\section{Attack}
\label{sec:attack}
\subsection{Target Program}
\paragraph{Target Program}
The following program is the target of our attack, it uses a command line argument to provide the payload and \Verb+strcpy+ for the buffer overflow, overwriting the return address after the 8 Byte buffer. Using vulnerable input functions also works though.
\bltCode{vuln.c}{c}{The Target Program}{thetargetprogram}
\paragraph{Compilation}
We compile the target program with the following command. There are several important options given in this command. Most importantly the \bltInlineVerb{-fno-stack-protector} option disables stack canaries which would otherwise directly terminate the program when the canary is overwritten. The \Verb+-m32+ option compiles the binary as a 32 Bit executable, this makes the attack easier. The \Verb+-static+ option makes the binary statically linked. Without this option there are only 50 gadgets available, considering most of them are not useful for our attack it is practically impossible to perform the attack with just these gadgets. The \Verb+-static+ option includes the \Verb+libc+ library in the executable, increasing the gadget count to over 8000. However, it is possible to determine the address of the dynamically linked library at runtime and adding an offset for each gadget to this address. This has been described by Saif El-Sherei~\cite{el-sherei} but will not be further discussed in this paper
\bltCommand{compilation.sh}{The compliation command}{thecompilationcommand}
\subsection{Phases of developing the attack}
\paragraph{Phases}
The attack consists of several phases
\begin{enumerate}
  \item Specify concrete goal with required program state and instructions
  \item Generate desired list of instructions and arguments (abstract payload)
  \item Extract gadgets using tools
  \item Search gadgets for instructions
  \item Generate payload using the gadgets according to the the abstract payload while making sure gadgets dont interfere with our desired program state. This step can be done using Python which we will show in a later section~\cref{howtopack}
  \item Insert payload into target
\end{enumerate}
\paragraph{Goal and abstract payload}
After specifying the goal and possibly simplifying it we have to write a list of instructions and arguments that achieve the goal, for this its favorable to directly use the format of the final payload except for using instructions instead of addresses as this will then allow to simply insert the found gadgets into this abstract payload.
\paragraph{Extract and search gadgets}
After extracting the gadgets using one of the above mentioned methods we can search for gadgets
\paragraph{struct.pack}
\Verb+struct.pack+ is a Python function that allows to easily generate our desired payload from the raw bytes. Bash then allows to directly pipe the generated payload into our target. In order to generate the payload we first have to fill the buffer and override the EBP with arbitary values as seen in line 2~\cref{howtopack}. This is usually done using easily recognizable characters, using the letter \Verb+A+ for this is common. It has the hex value \Verb+0x41+, doing this allows then to spot the buffer in a debugger like \Verb+gdb+. So in this example we fill the buffer with 8 \Verb+A+'s and 4 \Verb+B+'s. After that it is time to insert the addresses of the gadgets and the arguments. This is done by calling pack with the double word (64 Bit) while specifying the endianness, converting that to a string and adding it to the string as seen in line 3~\cref{howtopack}. After the whole payload has been generated we can print it and use the output directly for running the buffer overflow attack as mentioned above.
\bltCode{pack.py}{python}{How to use struct.pack}{howtopack}

@@

\section{Discussion}
Sources:

\bltInlineVerb{https://www.exploit-db.com/docs/english/28479-return-oriented-programming-(rop-ftw).pdf}
\bltInlineVerb{https://guyinatuxedo.github.io/5.1-mitigation_aslr_pie/index.html}
%%%% 8. BILBIOGRAPHY %%%%
\bibliographystyle{alpha}
\bibliography{abbrev3,crypto,biblio}
%%%% NOTES
% - Download abbrev3.bib and crypto.bib from https://cryptobib.di.ens.fr/
% - Use bilbio.bib for additional references not in the cryptobib database.
%   If possible, take them from DBLP.
\end{document}
